---
title: "Problem Set #1"
subtitle: "BST 258: Causal Inference -- Theory and Practice"
author: "Jenna Landy"
date: "02/16/2024"
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fontsize: 11pt
    geometry:
      - margin=1in
      - heightrounded
    number-sections: false
    colorlinks: true
    link-citations: true
    callout-appearance: simple
    callout-icon: false
    # figure options
    fig-width: 6
    fig-asp: 0.618
    fig-cap-location: bottom
    # code block options
    code-line-numbers: false
    code-block-bg: false
    highlight-style: nord
bibliography: refs.bib
---

{{< pagebreak >}}

## Question 1

::: {.callout-note title="Answer"}
My GitHub repo is here <https://github.com/jennalandy/bst258_psets>
:::

{{< pagebreak >}}

## Question 2

Consider a completely randomized experiment (CRE) with $ğ‘– = 1, ... , ğ‘›$
units, $ğ‘š$ of which are treated,\
where $ğ´_ğ‘–$ is 1 when unit $ğ‘–$ is treated.

a.  What is the marginal distribution of the treatment indicator $A$?

    ::: {.callout-note title="Answer"}
    We can view treatments in a completely randomized experiment as
    marbles of two colors (m green for treated, n-m red for untreated)
    in an urn. In this setting we know the marginal probability of a
    unit being treated is $m/n$.\
    \
    $A \sim Bernoulli(p)$ where $p = m/n$.
    :::

b.  What is the joint distribution of $A_i$ and $A_j$ for two units
    $i \ne j$? (Hint: This amounts to completing a contingency table.)

    ::: {.callout-note title="Answer"}
    Continuing with the idea of colored marbles in an urn, once a single
    marble is drawn (without replacement), it changes the probabilities
    for future draws. For example,
    $Pr(A_i = 1|A_j = 1) = \frac{m-1}{n-1}$.

    \
    We can compute a joint probability as
    $Pr(A_i = 1, A_j = 1) = Pr(A_j = 1) Pr(A_i = 1|A_j = 1) = \frac{m}{n}\cdot\frac{m-1}{n-1}$.\
    \
    Below is the contingency table representing the joint distribution
    of any two treatment assignments $A_i$ and $A_j$ where $i\ne j$.

    | $\mathbf{p(A_i, A_j) = }$ | $A_i = 1$                         | $A_i = 0$                             |
    |--------------------|-------------------------|----------------------------|
    | $A_j = 1$                 | $\frac{m}{n}\cdot\frac{m-1}{n-1}$ | $\frac{m}{n}\cdot\frac{n-m}{n-1}$     |
    | $A_j = 0$                 | $\frac{n-m}{n}\cdot\frac{m}{n-1}$ | $\frac{n-m}{n}\cdot\frac{n-m-1}{n-1}$ |

    Note that the off-diagonal probabilities are equivalent.
    :::

c.  What are $Var(A_i)$ and $Cov(A_i, A_j)$ for $i \ne j$?

    ::: {.callout-note title="Answer"}
    $Var(A_i) = \frac{m}{n}(1 - \frac{m}{n})$ using the fact that
    $A_i \sim Bernoulli(\frac{m}{n})$

    $Cov(A_i, A_j) = E[A_iA_j] - E[A_i]E[A_j] = \frac{m}{n}\cdot\frac{m-1}{n-1} - (\frac{m}{n})^2$

    **\
    Derivation:**

    -   $E[A_i] = \frac{m}{n}$ for all $i$ using the fact that
        $A_i \sim Bernoulli(\frac{m}{n})$

    -   $E[A_iA_j] = \sum_{a_i = 0}^1\sum_{a_j = 0}^1a_ia_jPr(A_i = a_i, A_j = a_j) = 1\cdot 1\cdot Pr(A_i = 1, A_j = 1)$
        because any term with $a_i = 0$ or $a_j = 0$ drops out. So
        $E[A_iA_j] = \frac{m}{n}\cdot\frac{m-1}{n-1}$.
    :::

d.  The sample Average Treatment Effect on the Treated (ATT) is
    $\theta_{ATT} = \frac{1}{m}\sum_{i = 1}^n A_i (Y_i(1) - Y_i(0))$.
    What is the sample ATT in expectation?

    ::: {.callout-note title="Answer"}
    $E[\theta_{ATT}] = \theta^{SATE}_n = \psi^{SATE}$ which is the
    sample average treatment effect.

    \
    **Derivation**:

    -   $E[A_i(Y_i(1) - Y_i(0))] = E[A_i]\cdot E[(Y_i(1) - Y_i(0))]$
        because randomization gives $A \perp Y(a))$.

    -   $E[A_i] = \frac{m}{n}$ as shown in (d).

    -   So
        $E[A_i(Y_i(1) - Y_i(0))] = \frac{m}{n}\cdot E[(Y_i(1) - Y_i(0))]$

    -   Then
        $E[\theta_{ATT}] = \frac{1}{m}\sum_{i = 1}^nE[A_i(Y_i(1) - Y_i(0))] = \frac{1}{n}\sum_{i = 1}^nE[Y_i(1) - Y_i(0)]$
        which we recognize as the sample average treatment effect
        $\theta^{SATE}_n = \psi^{SATE}$.
    :::

{{< pagebreak >}}

## Question 3

Consider an additive treatment effect model, i.e.,
$\theta = Y_i(1) âˆ’ Y_i(0)$ for all $i$, so $ğ‘Œ_ğ‘–(1) = ğ‘Œ_ğ‘–(0) + \theta$.
Show that $Var(ğ‘Œ_ğ‘–(1)) = Var(ğ‘Œ_ğ‘–(0))$ and that the correlation
$\rho(ğ‘Œ_ğ‘–(1), ğ‘Œ_i(0)) = 1$, where expectations are sample expectations,
i.e., $E[Y_ğ‘–(1)] â‰” \frac{1}{n}\sum_{i = 1}^nY_i(1)$.

::: {.callout-note title="Answer"}
To show, $Var(ğ‘Œ_ğ‘–(1)) = Var(ğ‘Œ_ğ‘–(0))$ I rely on the fact that $\theta$ is
a constant so $Var(\theta) = 1$. For both pieces, I use the fact that
$Cov(Y_i(0), \theta) = 0$, also because $\theta$ is a constant.

$$
\begin{aligned}
Y_i(1) &= Y_i(0) + \theta, \text{ so}\\
Var(Y_i(1)) & = Var(Y_i(0)) + Var(\theta) + Cov(Y_i(0), \theta)\\
&= Var(Y_i(0)) \checkmark
\end{aligned}
$$

$$
\begin{aligned}
Cov(Y_i(1), Y_i(0)) & = Cov(Y_i(0) + \theta, Y_i(0))\\
& = Cov(Y_i(0)_, Y_i(0)) + Cov(\theta, Y_i(0))\\
& = Var(Y_i(0)) + 0\\
\rho(Y_i(1), Y_i(0)) & = \frac{Cov(Y_i(1), Y_i(0))}{\sqrt{Var(Y_i(1))\cdot Var(Y_i(0))}}\\
& = \frac{Var(Y_i(0))}{\sqrt{Var(Y_i(0))^2}} = 1\checkmark
\end{aligned}
$$
:::

{{< pagebreak >}}

## Question 4

It's tea time: (a hologram of) R.A. Fisher places eight cups of tea
(with milk) in front of you and\
asks you to identify which cups had tea poured before milk and
vice-versa. Prior to giving you\
the cups of tea, Fisher poured milk before tea in four of them and tea
before milk in the other\
four. The ordering in which the cups have been served is random. What is
the probability that\
you correctly guess 0, 1, 2, 3, or 4 of all of the cups that had tea
poured first?

::: callout-note
Guessing at completely random, then guessing which four cups had tea
poured first is equivalent to partitioning the eight glasses into four
that are "guessed" and four that are not. We can compute the probability
of "x" correct guesses by considering the number of partitions with "x"
correct divided by the total number of partitions.\

There are ${8 \choose 4} = 70$ total possible partitions.

-   **0 successes**: To have 0 successful guesses, we must guess 0 of
    the 4 "tea-first" cups and 4 of the 4 "milk-first" cups. This is
    ${4\choose 4}\cdot {4\choose 0} = 1$ partition that will lead to 0
    successes, giving $Pr(0 \text{ successes}) = 1/70 \approx 0.014$

-   **1 success**: There are ${4 \choose 1}{4\choose 3} = 16$ partitions
    that will lead to 1 success, giving
    $Pr(1 success) = 16/70 \approx 0.229$

Continuing this pattern gives

-   **2 successes**: $36/70 = 0.514$

-   **3 successes**: $16/70 = 0.229$

-   **4 successes**: $1/70=0.014$
:::

## Question 5

The table below displays the success rates of two distinct,
investigational treatments for kidney\
stones, labeled as $ğ´$ and $ğµ$. A study enrolls $ğ‘› = 700$ participants,
assigning $ğ‘› = 350$ individuals\
to either of the treatment arms, $ğ´$ and $ğµ$. To summarize data from the
study, individuals' kidney\
stones are categorized as either small or large, and Table 1 is
constructed to summarize the success\
rates of each of the two treatments.

|              | Treatment A   | Treatment B   |
|--------------|---------------|---------------|
| Small Stones | 93% (81/87)   | 87% (234/270) |
| Large Stones | 73% (192/263) | 69% (55/80)   |
| Both         | 78% (273/350) | 83% (289/350) |

: Table 1: Success rates in arms ğ´ and ğµ versus kidney stone size

Studying Table 1, your colleague remarks at the discrepancy between the
superior overall success rate of treatment $ğµ$ and its relatively lower
success (versus treatment $ğ´$) when stratifying cases by\
kidney stone size.

a.  Describe possible factors that might have contributed to this
    seemingly contradictory result.

    ::: {.callout-note title="Answer"}
    We were not told that the two treatments were randomly assigned. In
    fact, we see that treatment $A$ is assigned to 76.7% of patients
    with large stones ($263/(263+80)$) but only to 24.4% of patients
    with small stones ($87/(97 + 270)$). Further, large stones are
    generally less likely to be treated successful (72% overall success
    for large versus 88% overall success for small).\

    It is likely that patients with more extreme (large) kidney stones
    were assigned treatment $A$ because doctors believe it is most
    effective in these cases. Perhaps treatment $A$ is more invasive and
    only necessary for large kidney stones.\

    This is an example of Simpson's paradox. The size of the patient's
    stones is a confounder of the association between treatment and
    success.
    :::

b.  Alarmed by this discrepancy, your colleague asks you to further
    segment the results by reported gender. This newly refined look at
    the data suggests that for both small and large\
    kidney stones, treatment $ğµ$ is consistently more effective than
    treatment $ğ´$ across all genders. Construct a hypothetical (i.e.,
    candidate) table that illustrates this (ensure that your\
    candidate table is consistent with the information given in Table
    1).

    ::: {.callout-note title="Answer"}
    The expanded table below shows that for both small and large kidney
    stones, treatment B is more effective than treatment A across all
    genders.

    |                  | Treatment A     | Treatment B     |
    |------------------|-----------------|-----------------|
    | **S**mall Stones |                 |                 |
    | Female           | 70.0% (7/10)    | 78.6% (118/150) |
    | Male             | 96.1% (74/77)   | 96.6% (116/120) |
    | Large Stones     |                 |                 |
    | Female           | 65% (65/100)    | 67.1% (47/70)   |
    | Male             | 79.4% (127/163) | 80% (8/10)      |
    | **Total**        | 273/350         | 289/350         |
    :::

c.  What is this phenomenon (it has a name)? What are its broader
    implications and significance in the interpretation of data?

    ::: {.callout-note title="Answer"}
    This is an example of Simpson's paradox. The broader implication of
    Simpson's paradox is that observed population-level associations
    between two variables, $A$ and $Y$, may be due to mutual association
    with other (possibly unmeasured) variables, known as confounders,
    $L$. This can be seen as population-level associations that go away
    or reverse once the data is divided into sub-populations.\

    The key takeaway is that *correlation does not imply causation*, and
    there may always be unmeasured confounders in observational data.
    :::

## Question 6

::: callout-warning
Below are helper functions to compute our difference in means estimates
for the ATE, the variance of this estimator as defined in lecture notes,
and to simulate the sharp null distribution.

```{r}
# Computes difference in means estimate for ATE
compute_ATE_DM <- function(treatments, outcomes) {
    mean(outcomes$Y_1[treatments == 1]) - 
    mean(outcomes$Y_0[treatments == 0])
}

# Computes upper limit of variance of DM estimator for ATE
compute_Var_ATE_DM <- function(treatments, outcomes) {
    n1 <- sum(treatments == 1)
    n0 <- sum(treatments == 0)
    
    Y0s <- outcomes$Y_1[treatments == 0]
    Y1s <- outcomes$Y_1[treatments == 1]
    v0 <- sum((Y0s - mean(Y0s))**2)/(n0-1)
    v1 <- sum((Y1s - mean(Y1s))**2)/(n1-1)

    v1/n1 + v0/n0
}

# Simulate Fisher's sharp null distribution
# Y_i(1) - Y_i(0) = 0, set Y_i(1) and Y_i(0) 
#     to same value for all repetitions 
simulate_sharp_null <- function(
    N, outcomes,
    repetitions = 10000
) {
    outcomes$Y_1 = outcomes$Y_0
    ATE_DM_null <- vector(length = repetitions)
    for (rep in 1:repetitions) {
        treatments = sample(c(
            rep(0, round(N*pr_treatment)), 
            rep(1, N - round(N*pr_treatment))
        ))
        ATE_DM_null[rep] = compute_ATE_DM(treatments, outcomes)
    }
    return(ATE_DM_null)
}
```

Below is a function to compute the power of each test given sample size
$N$

```{r}
compute_powers <- function(
    N, mu_1, mu_0, sigmasq,
    alpha, nsim, pr_treatment
) {
    Y_1 = rnorm(N, mu_1, sd = sqrt(sigmasq))
    Y_0 = rnorm(N, mu_0, sd = sqrt(sigmasq))
    outcomes = list(Y_1 = Y_1, Y_0 = Y_0)
    
    weak_null_cutoff = qt(c(alpha/2, 1-alpha/2), df = N-2)
    
    sharp_null = simulate_sharp_null(N, outcomes)
    sharp_null_cutoff = quantile(sharp_null, c(alpha/2, 1-alpha/2))
    
    ATE_DM = vector(length = nsim)
    Var_ATE_DM = vector(length = nsim)
    for (sim in 1:nsim) {
        treatments = sample(c(
            rep(0, round(N*pr_treatment)), 
            rep(1, N - round(N*pr_treatment))
        ))
        ATE_DM[sim] = compute_ATE_DM(treatments, outcomes)
        Var_ATE_DM[sim] = compute_Var_ATE_DM(treatments, outcomes)
    }
        
    # power = Pr(reject)
    reject_weak_null <- ATE_DM/sqrt(Var_ATE_DM) < weak_null_cutoff[1] | 
                        ATE_DM/sqrt(Var_ATE_DM) > weak_null_cutoff[2]
    
    reject_strong_null <- ATE_DM < sharp_null_cutoff[1] | 
                          ATE_DM > sharp_null_cutoff[2]
    
    weak_power = mean(reject_weak_null)
    strong_power = mean(reject_strong_null)
    
    return(list(
        N = N,
        weak_norm = weak_power,
        strong = strong_power
    ))
}
```

This code runs the power analysis for each sample size N in
`{20, 50, 100, 200, 500}` and records the power for each test in a
table.

```{r}
set.seed(123)

mu_0 = 0
mu_1 = 1/10
sigmasq = 1/16
alpha = 0.05
nsim = 1000
pr_treatment = 0.5

powers = matrix(ncol = 3, nrow = 0)
for (N in c(20, 50, 100, 200, 500)) {
    powers = rbind(
        powers, 
        compute_powers(
            N = N,
            mu_1 = mu_1, mu_0 = mu_0, sigmasq = sigmasq,
            alpha = alpha, nsim = nsim, pr_treatment = pr_treatment
        )
    )
}

powers
```

```{r}
plot(
    powers[,1], powers[,2], col = 'red', 
    xlab = "sample size (n)", ylab = "Power", 
    main = "Power for Tests of the Weak and Strong Nulls"
)
points(powers[,1], powers[,3], col = 'blue')
legend(
    "bottomright", legend = c('weak','strong'), 
    col = c('red','blue'), pch = 1
)
```

We see the expected trend that power increases with sample size for both
tests. However, we see that the power of the strong null is not always
greater than the power of the weak null. In fact, which the test that is
more powerful changes at each sample size without a clear pattern.\

This is because although Neymon's weak null implies Fisher's strong
null, that doesn't mean that Neymon's rejection region is *not* a subset
of Fisher's rejection region. This is because there. My intuition says
this is because Fisher's strong null is not in terms of the ATE, so it
doesn't make sense to put them on the same axes or scale.
:::

## Question 7

Consider a completely randomized experiment (CRE) having enrolled
$ğ‘– = 1, ... , ğ‘›$ study units, $ğ‘š$ of which receive the treatment
condition. Let $ğ´_ğ‘–$ be the indicator of the $i$th unit having received
the treatment, and, further, define
$\bar{Y}_1 = \frac{1}{m}\sum_{i = 1}^nA_iY_i$ be the average outcome of
the treated units, and similarly define
$\bar{Y}_0 = \frac{1}{n-m}\sum_{i = 1}^n (1-A_i)Y_i$. You have been
tasked with estimating the average treatment effect (ATE), for which it
suï¬€ices to solve the following least squares program

$$min_{\alpha,\beta}\frac{1}{2n}\sum_{i = 1}^n(Y_i - \alpha - \beta A_i)^2$$

a.  Solve the linear program in $(\alpha, \beta)$ to obtain solutions
    for each of the two parameters, denoting these
    $(\hat \alpha, \hat \beta)$.

    ::: {.callout-note title="Answer"}
    $(\hat \alpha, \hat \beta) = (\bar{Y}_0, \bar{Y}_1 - \bar{Y}_0)$

    **\
    Derivation**:

    -   $f = \frac{1}{2n}\sum_{i = 1}^n(Y_i - \alpha - \beta A_i)^2$

    -   Set the partial derivatives to zero

        $\frac{\partial}{\partial \alpha}f = -\frac{1}{n}\sum_{i = 1}^n (Y_i - \alpha - \beta A_i) \stackrel{set}{=} 0 \implies \beta m = \sum_{i = 1}^nY_i - n\alpha$

        $\frac{\partial}{\partial \beta}f = -\frac{1}{n}\sum_{i = 1}^n A_i(Y_i - \alpha - \beta A_i) \stackrel{set}{=} 0 \implies \beta m = \sum_{i = 1}^nA_iY_i - m\alpha$

    -   Solve for $\alpha$ using the above system of equations

        $\sum_{i = 1}^nY_i - n\alpha = \sum_{i = 1}^nA_iY_i - m\alpha$

        $\sum_{i = 1}^nY_i - \sum_{i = 1}^nA_i Y_i = \alpha (n - m)$

        $\hat \alpha = \frac{1}{n-m}\sum_{i = 1}^nY_i(1-A_i) = \bar{Y}_0$

    -   Plug $\hat{\alpha}$ into the
        $\beta m = \sum_{i = 1}^nA_iY_i - m\alpha$ and solve for $\beta$

        $\beta m = \sum_{i = 1}^nA_iY_i - m\bar{Y}_0 = m\bar{Y}_1 - m\bar{Y}_0$

        $\hat{\beta} = \bar{Y}_1 - \bar{Y}_0$

    -   Check that this is in fact a minimum

        $\frac{\partial^2}{\partial \alpha^2}f = 1$

        $\frac{\partial^2}{\partial \beta^2} f = -\frac{1}{n}\sum_{i = 1}^n -A_i^2 = \frac{1}{n}\sum_{i = 1}^n A_i = m/n$

        $\frac{\partial^2}{\partial\alpha \partial \beta} = \frac{\partial^2}{\partial \beta \partial \alpha} = -\frac{1}{n}\sum_{i = 1}^n- A_i = m/n$

        $H = \begin{bmatrix}1 & m/n\\ m/n & m/n\end{bmatrix}$

        $det(H) = m/n -(m/n)^2 >0$ because $m/n <1$ so $(m/n)^2 < m/n$
        AND both $\frac{\partial^2}{\partial \alpha^2}f>0$ and
        $\frac{\partial^2}{\partial \beta^2}f >0$. Together, this shows
        that the estimates $\hat \alpha$ and $\hat \beta$ are in fact
        minimums of the loss $f$.
    :::

b.  Is $\hat{\beta}$ a valid estimator of the ATE? Explain your answer.

    ::: {.callout-note title="Answer"}
    The estimate $\hat \beta$ is equivalent to our standard difference
    ofmeans estimator. We showed in class that in the case of a
    completely randomized experiment, this estimator is unbiased for the
    SATE assuming consistency.

    \
    \
    It we were instead looking at a Bernoulli trails experiment, it
    would beunbiased for the ATE. Because this is not the case,
    $\hat \beta$ is not an appropriate estimator of the ATE.
    :::

## References

::: {#refs}
:::
